\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-07 k-planar.tex

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.  We say that a graph $G$ is \emph{contained in} a graph $X$ if $G$ is isomorphic to a subgraph of $X$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (improved by \citet{UWY}).

\begin{thm}[\citep{DJMMUW20,UWY}]
	\label{PlanarProduct}
	Every planar graph is a subgraph of:
	\begin{compactenum}[(a)]
		\item $H\boxtimes P$ for some graph $H$ of treewidth at most $8$ and for some path $P$,
		\item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

@@

% \note{DW}{I suggest we replace $8$ by $6$ here, and replace ``The starting point for this work is the following `product structure theorem' for planar graphs by \citet{DJMMUW20} (improved by \citet{UWY})''. Okay?}

@@

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
      \includegraphics{figs/one_planar_example-1} &
      \includegraphics{figs/one_planar_example-2} \\
      (a) & (b) \\[1em]
      \includegraphics{figs/one_planar_example-3} &
      \includegraphics{figs/one_planar_example-4} \\
      (c) & (d) \\[1em]
    \end{tabular}
  \end{center}
  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.} 
  \label{one_planar_example} 
  \note{PM}{I suggest we only keep parts (c) and (d) of this figure.}
\end{figure}

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

>>>>

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.  We say that a graph $G$ is \defin{contained in} a graph $X$ if $G$ is isomorphic to a subgraph of $X$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (improved by \citet{UWY}).

\begin{thm}[\citep{DJMMUW20,UWY}]
	\label{PlanarProduct}
	Every planar graph is contained in: 
	%a subgraph of:
	\begin{compactenum}[(a)]
		\item $H\boxtimes P$ for some graph $H$ of treewidth at most $8$ and for some path $P$,
		\item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

@@

 \note{DW}{replace $8$ by $6$ and update elsewhere}

@@

%Let $G$ be an edge-maximal 1-plane multigraph with no two parallel edges on the boundary of a single face.  Here, edge-maximal should be taken to mean that, if any two vertices $v$ and $w$ appear on a common face\footnote{The \defin{faces} of an embedded graph $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.} $F$, then there is an edge $vw\in E(G)$ that is contained in the boundary of $F$.  \note{DW}{Each face of $G$ is one of two types, either bounded by three edges (none of which are crossed), or bounded by one edge that is not crossed plus portions of two crossing edges. Right? If this is true, we should say it here, which would help to clear up the confusion below. Add a figure.} 
%We assume that no two edges incident to a common vertex cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}
%
%\note{DW}{I suggest we say that it follows from Euler's formula that $G$ is finite. (To see this, say $G$ has $n$ vertices, $m$ non-crossed edges, and $k$ crossings. Let $G'$ be the plane graph obtained from $G$ by adding a dummy vertex at each crossing. So $G'$ is a plane multigraph in which every face is bounded by a 3-cycle. It follows from Euler's formula that $m+4k = |E(G')| = 3( |V(G')|-2) = 3 ( n+k-2)$, implying	$m+k = 3 (n-2)$ and $|E(G)| = m+2k \leq 6(n-2)$, which can also be shown by deleting one edge from each crossing point. ) }
%
%\note{DW}{It seems to be that $G$ can be obtained from a plane multigraph with faces of size 3 or 4 such that no edge is in two 3-faces, by adding a pair of crossing edges across each 4-face. I think this characterises edge-maximal 1-planar multigraphs with no 2-faces. Right? This should be stated as a lemma. Does this viewpoint help? Is it known?}
%
%\note{DW}{Say $G$ is obtained from a plane multigraph by making each face a clique. Is $G\subseteq H \boxtimes P \boxtimes K_{f(\omega(G)}$? for some treewidth 3 graph $H$.}
%
%A \defin{kite} in $G$ is the subgraph $K=G[\{v,w,x,y\}]$ induced by the endpoints of a pair of crossing edges $vw,xy\in E(G)$.  It follows from edge-maximality that every kite is isomorphic to the complete graph $K_4$. \note{DW}{We need to define a kite to be the subgraph with vertex-set $\{v,w,x,y\}$ and edge-set $\{vx,xy,vx,vy,wy,wx\}$ where $vx,vy,wy,wx$ are the edges on the `outside' of the kite-faces. Then it really is a simple $K_4$. I think this is what we intended. We should add a figure showing a kite and kite-faces. }
%The edges $vw$ and $xy$ are called \defin{spars} of $K$.  The cycle $vxwy$ is called the \defin{sail} of $K$.  It follows from edge-maximality that none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of $G$. 
%
%\referee{2}{Page 13. Perhaps I am misunderstanding something, but I do not see why
%'none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of
%$G$.'  For example, see the attached PDF for a picture where $vw$, $xw$, $wy$,
%and $yv$ are all crossed by other edges of $G$.\\
%\includegraphics[width=60mm]{kite}}
%\note{DW}{We need to explain more about how edge-maximality fixes this problem. In this case we can draw other $xv$, $vy$, $yw$ $wx$ edges close to the central crossing edges.} Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$.
%
%\referee{2}{Page 13. The attached PDF also shows that the definition of 'kite
%face' may not be well-defined.  For example, I do not see why there
%cannot be some edges of $G$ 'inside' a kite face.  Even if this is not
%the case, I think it is more precise to say that a kite face has 'two
%and a half edges' and two vertices of $G$ on its boundary' rather than
%'three edges and two vertices of $G$ on its boundary'.}

\note{DW}{This should be fixed if we define kite more precisely. A kite-face is bounded by one edge and two half-edges right?}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
%      \includegraphics{figs/one_planar_example-1} &
%      \includegraphics{figs/one_planar_example-2} \\
      (a) & (b) \\[1em]
%      \includegraphics{figs/one_planar_example-3} &
%      \includegraphics{figs/one_planar_example-4} \\
      (c) & (d) \\[1em]
    \end{tabular}
  \end{center}
  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; (b)~an edge-maximal $1$-plane graph $G$; (c)~the kites in $G$; (d)~a plane multigraph $G'$ obtained by removing one spar from each kite of $G$.} 
  \label{one_planar_example} 
  \note{PM}{I suggest we only keep parts (c) and (d) of this figure.}
\end{figure}

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup (since the new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$)?. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}
